### 2021-06-08 papers/thesis/full/abstract/abstract.tex
Addressed most comments so far.

We move on working in 2D, where in recent years sparse spectral methods for solving PDEs have been derived using hierarchies of classical orthogonal polynomials on intervals, disks, and triangles. Presenting a new framework for choosing a suitable orthogonal polynomial basis for more general 2D domains defined via an algebraic curve as a boundary, this work builds on the observation that sparsity is guaranteed due to this definition of the boundary, and that the entries of partial differential operators can be determined using formulae in terms of (non-classical) univariate orthogonal polynomials. Triangles and the full disk are then special cases of our framework, which we formalise for the disk-slice and trapezium cases.

>>>>

We move on to working in 2D, where in recent years sparse spectral methods for solving PDEs have been derived using hierarchies of classical orthogonal polynomials on intervals, disks, and triangles. Presenting a new framework for choosing a suitable orthogonal polynomial basis for more general 2D domains defined via an algebraic curve as a boundary, this work builds on the observation that sparsity is guaranteed due to this definition of the boundary, and that the entries of partial differential operators can be determined using formulae in terms of (non-classical) univariate orthogonal polynomials. Triangles and the full disk are then special cases of our framework, which we formalise for the disk-slice and trapezium cases.

### 2021-06-08 papers/thesis/full/acknowledgements/acknowledgements.tex
Pushing the pdf of thesis
Addressed most comments so far.

I would like to thank the MPE CDT staff for any knowledge and wisdom they have bestowed, and in particular say a huge thank you to Colin Cotter for his guidance, understanding and general enthusiasm (it meant a lot). And to all my fellow 'Charlies' -- I truly mean it when I say you are all such great, smart and kind people, and you all deserve the absolute best for the future.

I thank my family and friends for all their support and words of encouragement -- it all helped a lot. An extra thanks to all my undergrad mates for many well needed fun times too.

A special thanks goes to my partner Jessica for putting up with me and helping me when things got tough during my 4 years at Imperial College London.

@@

I would like to express (whatever feelings I have) to:

\begin{itemize}
 \item My supervisor
 \vspace*{3mm}
 \item My second supervisor
 \vspace*{3mm}
 \item Other researchers
 \vspace*{3mm}
 \item My family and friends
\end{itemize}

>>>>

I would like to thank the MPE CDT staff for any knowledge and wisdom they have bestowed, and in particular say a huge thank you to Colin Cotter for his guidance, understanding and general enthusiasm (it meant a lot). And to all my fellow \enquote{Charlies} -- I truly mean it when I say you are all such great, smart and kind people, and you all deserve the absolute best for the future.

I thank my family and friends for all their support and words of encouragement -- it all helped a lot.

A special thanks goes to my partner Jess for putting up with me and helping me when things got tough during my 4 years at Imperial College London.

@@

%I would like to express (whatever feelings I have) to:
%
%\begin{itemize}
% \item My supervisor
% \vspace*{3mm}
% \item My second supervisor
% \vspace*{3mm}
% \item Other researchers
% \vspace*{3mm}
% \item My family and friends
%\end{itemize}

### 2021-06-08 papers/thesis/full/ch1-introduction/introduction.tex
Addressed most comments so far.

\chapter{Introduction}

@@

One could say this is surprising, given that there is a long history of around 150 years associated with multivariate OPs, beginning with Hermite first presenting the multivariate Hermite polynomials in 1865 \cite{appel1926fonctions, ismail2017review}. Zernike polynomials \cite{zernike1934diffraction} were first introduced in 1934, as another example, that are a group of bivariate polynomials orthogonal on a unit circle. Koornwinder in 1975 described a method for constructing two-variable OPs from univariate OPs \cite{koornwinder1975two}. However, few books have been published over the years on the topic of multivariate orthogonal polynomials \cite{dunkl2014orthogonal}. This branch of mathematics has an encouraging future though, not least as a basis for sparse spectral methods for solving partial differential equations (PDEs) on multidimensional domains. Spectral methods have been developed for solving PDEs as an alternative to finite difference and finite element methods. For example, in recent years spectral methods have been established on the triangle \cite{olver2019triangle} using OPs inspired by the Koornwinder approach for defining them, and on the disk \cite{vasil2016tensor} using Zernike polynomials.

@@

The motivation for this project was to help address this problem while still utilising a spectral approach. More precisely, we aim to develop a sparse spectral method for solving PDEs on the spherical cap as a surface in 3D, with a simple extension to a spherical band. Together, these frameworks can be pieced together to create a spectral element method for the whole sphere, or further developed to investigate spectral methods on other spherical subdomains. This, however, is future work beyond the scope of this thesis. By spectral element method, we mean a finite element method (FEM) that uses high degree basis polynomials for its elements (this could also be referred to as a $p$-FEM with large $p$). In other words, we can use our spectral methods developed for the spherical band and cap elements as part of a finite element framework. By using this approach, one would be able to avoid having to complete the global spectral transform (in particular, the global Legendre transform) and instead be able to simply apply the local element transforms in parallel. Moreover, by still using a spectral approach, one can maintain the high accuracy and excellent error properties that such methods bring. There are also potential applications in physics too, particularly in astrophysics, where solving PDEs on the sphere surface and working in spherical geometries is also desirable (e.g. \cite{vasil2019tensor, reinecke2013libsharp, beyer2014numerical, varshalovich1988quantum, slevinsky2018spectral, rubinstein2015scalar}).

@@

Chapter 2 of this thesis provides an introduction to sparse spectral methods via the spherical harmonics on the whole sphere surface. Here, we think of the spherical harmonics in a non-traditional way and write them as a group of multidimensional orthogonal polynomials in $(x,y,z)$ as opposed to functions of spherical coordinates. We present an OP framework for how the spherical harmonics can be used to expand functions defined on the sphere as multidimensional polynomials in $x, y, z$, and how differential operators can be applied as banded-block-banded matrix operators to coefficient vectors for a function's expansion. Further, we demonstrate how the vector spherical harmonics can be used as an orthogonal basis for vector valued functions lying in the \enquote{tangent bundle} of the sphere, and thus how one can additionally derive gradient and divergence operators.

In Chapter 3 we move on to working in 2D, where in recent years sparse spectral methods for solving PDEs have been derived using hierarchies of classical orthogonal polynomials on intervals, disks, and triangles. Presenting a new framework for choosing a suitable orthogonal polynomial basis for more general 2D domains defined via an algebraic curve as a boundary, this work builds on the observation that sparsity is guaranteed due to this definition of the boundary, and that the entries of partial differential operators can be determined using formulae in terms of (non-classical) univariate orthogonal polynomials, which we define. Triangles and the full disk are then special cases of our framework, which we formalise for the disk-slice and trapezium cases.

With a greater knowledge base in our quiver, we can adapt the techniques learnt from the founding of the disk-slice formulation to surfaces in 3D in Chapter 4. Using the same family of (non-classical) 1D OPs, we present a suitable orthogonal polynomial basis for the spherical cap, a subdomain of the surface of a unit sphere, complete with sparse differential operators. A relatively simple adaption permits this framework to be extended to a spherical band. From here, a spectral element method could be devised for the whole sphere using the aforementioned as elements.

Finally, we summarise the work we have presented and detail avenues for future directions that one could take it in Chapter 5.

>>>>

\chapter{Introduction}\label{CHAPTER:intro}

@@

One could say this is surprising, given that there is a long history of around 150 years associated with multivariate OPs, beginning with Hermite first presenting the multivariate Hermite polynomials in 1865 \cite{appel1926fonctions, ismail2017review}. Zernike polynomials \cite{zernike1934diffraction} were first introduced in 1934, as another example, that are a group of bivariate polynomials orthogonal on a unit circle. Koornwinder in 1975 described a method for constructing two-variable OPs from univariate OPs \cite{koornwinder1975two}. However, few books have been published over the years on the topic of multivariate orthogonal polynomials, with the notable exception of the book by Dunkl \& Xu \cite{dunkl2014orthogonal}. As Dumitriu, Edelman and Shuman said in 2007, \enquote{\textit{[multivariate orthogonal polynomials] are understudied, underapplied, and important applications may be being missed}} \cite{dumitriu2007mops}, while in 2016, Olver opined that \enquote{\textit{there is still a long way to go before multivariate orthogonal polynomials can reach their full potential in applications}}. He hypothesised that \enquote{\textit{[p]art of this neglect may be due to the fact that mutually orthogonal polynomials are not uniquely defined: there is no canonical ordering}} and continued \enquote{\textit{[e]ven when explicit constructions are known, these are often unwieldy}} \cite{olver2016review}.

This branch of mathematics has an encouraging future though, not least as a basis for sparse spectral methods for solving partial differential equations (PDEs) on multidimensional domains. Spectral methods have been developed for solving PDEs as an alternative to finite difference and finite element methods. For example, in recent years spectral methods have been established on the triangle \cite{olver2019triangle} using OPs inspired by the Koornwinder approach for defining them, and on the disk \cite{vasil2016tensor} using Zernike polynomials.

@@

The motivation for this project was to help address this problem while still utilising a spectral approach. More precisely, we aim to develop a sparse spectral method for solving PDEs on the spherical cap as a surface in 3D, with a simple extension to a spherical band. Together, these frameworks can be pieced together to create a spectral element method for the whole sphere, or further developed to investigate spectral methods on other spherical subdomains. This, however, is future work beyond the scope of this thesis. By spectral element method, we mean a finite element method (FEM) that uses high degree basis polynomials for its elements (this could also be referred to as a $p$-FEM with large $p$). In other words, we can use our spectral methods developed for the spherical band and cap elements as part of a finite element framework. By using this approach, one would be able to avoid having to complete the global spectral transform (in particular, the global Legendre transform) and instead be able to simply apply the local element transforms in parallel. Moreover, by still using a spectral approach, one can maintain the high accuracy and excellent error properties that such methods bring. Solving PDEs on a section of the sphere surface is still useful for weather prediction in its own right too. Moreover, there are also prospective applications in physics, particularly in astrophysics, where solving PDEs on the sphere surface and working in spherical geometries is also desirable (e.g. \cite{vasil2019tensor, reinecke2013libsharp, beyer2014numerical, varshalovich1988quantum, slevinsky2018spectral, rubinstein2015scalar}).

@@

\bsrefchapter{CHAPTER:sphericalharmonics} of this thesis provides an introduction to sparse spectral methods via the spherical harmonics on the whole sphere surface. Here, we think of the spherical harmonics in a non-traditional way and write them as a group of multidimensional orthogonal polynomials in $(x,y,z)$ as opposed to functions of spherical coordinates. We present an OP framework for how the spherical harmonics can be used to expand functions defined on the sphere as multidimensional polynomials in $x, y, z$, and how differential operators can be applied as banded-block-banded matrix operators to coefficient vectors for a function's expansion. Further, we demonstrate how the vector spherical harmonics can be used as an orthogonal basis for vector valued functions lying in the \enquote{tangent bundle} of the sphere, and thus how one can additionally derive gradient and divergence operators.

In \bsrefchapter{CHAPTER:diskslice} we move on to working in 2D, where in recent years sparse spectral methods for solving PDEs have been derived using hierarchies of classical orthogonal polynomials on intervals, disks, and triangles. Presenting a new framework for choosing a suitable orthogonal polynomial basis for more general 2D domains defined via an algebraic curve as a boundary, this work builds on the observation that sparsity is guaranteed due to this definition of the boundary, and that the entries of partial differential operators can be determined using formulae in terms of (non-classical) univariate orthogonal polynomials, which we define. Triangles and the full disk are then special cases of our framework, which we formalise for the disk-slice and trapezium cases. The work in this chapter has been previously published in \cite{snowball2019sparse}

With a greater knowledge base in our quiver, we can adapt the techniques learnt from the founding of the disk-slice formulation to surfaces in 3D in \bsrefchapter{CHAPTER:sphericalcaps}. Using the same family of (non-classical) 1D OPs, we present a suitable orthogonal polynomial basis for the spherical cap, a subdomain of the surface of a unit sphere, complete with sparse differential operators. A relatively simple adaption permits this framework to be extended to a spherical band. From here, a spectral element method could be devised for the whole sphere using the aforementioned as elements. The work in this chapter has been accepted for publication \cite{snowball2020sparse}.

Chapter 5 gives a summary of the work we have presented and details avenues for future directions that one could take it.

Finally, as an appendix, we outline how one could approach a vector OP basis for the \enquote{tangent bundle} of the spherical cap, analogous to the vector spherical harmonics on the whole sphere. This constitutes \bsrefappendix{appendix:tangent}.
